\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath, amssymb}
\usepackage{array}
\usepackage{multirow}
\usepackage{float}
\usepackage[hidelinks]{hyperref}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{lineno}
\linenumbers
\captionsetup{font=small,labelfont=bf}

\title{\textbf{Biological Continued Pretraining Reshapes the Capability\\
Profile of a Foundation Model Without Catastrophic Forgetting}}

\author{Liang Wang\thanks{Corresponding author. Email: wangliang.f@gmail.com}\\
\small School of Artificial Intelligence and Automation\\
\small Huazhong University of Science and Technology\\
\small Wuhan, Hubei 430070, P.R. China}
\date{July 2026}

\begin{document}
\maketitle

\noindent\textbf{Highlights}
\begin{itemize}\itemsep2pt
\item A training-free re-analysis of one 26B MoE lineage isolates the effect of biological
continued pretraining (CPT) from tokenizer changes and from fine-tuning.
\item Biological CPT does not cause catastrophic forgetting; it \emph{raises} general knowledge
(MMLU $+13$ pts) and code generation (MBPP pass@1 $0.33\!\to\!0.63$).
\item CPT also makes chain-of-thought reasoning 41\% shorter and near-backtrack-free while
preserving accuracy --- an effect invisible to accuracy metrics.
\item A consistent CPT-lifts / SFT-narrows division of labor recurs across four axes, reframing
biological sequence as structured scientific data that reshapes a model's capability profile.
\item A controlled seven-model experiment (varying only the CPT data mixture) confirms this
causally: raising biological share to 50\% preserves general ability (MMLU $-0.6$ pt) while
biology rises monotonically --- data mixture is a general-preserving lever on capability.
\item A small-model, matched-compute causal probe (GPT-2, 3 seeds) finds that biological CPT
\emph{reproducibly lifts} paraphrase detection (PAWS-en, $+3.1$ pt) --- the direct structural
counterpart of protein-homology transfer --- alongside a real cost to long-range discourse
modeling, an honestly reported partial, not uniform, transfer.
\end{itemize}

\bigskip
\noindent\fbox{\parbox{\linewidth}{%
\textbf{The Bigger Picture.}\\[2pt]
Adapting a general-purpose AI model to a specialized domain --- here, the language of DNA and
proteins --- is usually assumed to come at a cost: teach it biology and it forgets how to reason
about everything else. This ``no free lunch'' intuition shapes how practitioners budget compute
and whether they attempt domain adaptation at all. We test the assumption directly, and without
running any new training, by comparing three snapshots of the same model taken before and after
biological training. The result overturns the intuition: feeding the model raw biological
sequence made it \emph{better} at general knowledge, at writing code, and even changed \emph{how}
it reasons --- producing shorter, more decisive chains of thought without losing accuracy. The
gains appear during the sequence-pretraining stage and are partly given back during
task-specific fine-tuning, revealing that the two stages play complementary rather than
interchangeable roles. This suggests a broader principle for data-centric AI: structured
scientific data --- biological sequence today, and by extension code, mathematics, and
chemistry --- is not merely knowledge to be absorbed but a lever that reshapes what a foundation
model can do. A controlled follow-up, in which we retrain the model seven times while changing
only the data mixture, confirms the lever directly: more biological data buys more scientific
capability at almost no cost to general ability. A final, smaller-scale experiment asks the
question in reverse: does biological training feed anything back into ordinary language
understanding? Under a causal design that holds compute exactly fixed, it does --- reproducibly
improving a model's ability to detect when two sentences paraphrase each other --- though not
without cost elsewhere, a nuance we report honestly rather than smooth over.
}}
\bigskip

\begin{abstract}
\noindent It is widely assumed that continued pretraining (CPT) on a narrow, out-of-distribution
corpus such as raw biological sequence must trade away a general-purpose model's broad
competence --- the ``alignment tax'' or catastrophic-forgetting intuition. We test this
directly, without any new training, by re-analyzing three checkpoints from a single lineage
of a 26B-parameter Mixture-of-Experts model (Gemma-4-26B-A4B): the instruction-tuned base,
the same model after biological CPT (8.7B tokens of DNA, protein, and biomedical text), and
after subsequent supervised fine-tuning (SFT). Across three independent capability axes ---
general knowledge/reasoning (MMLU, ARC, HellaSwag), code generation (MBPP), and biomedical
knowledge (BixBench) --- we find that biological CPT does \emph{not} degrade the model; it
\textbf{lifts} it: MMLU $+13$ points, MBPP pass@1 nearly doubles ($0.33\!\to\!0.63$), and
BixBench discrimination rises sharply (MCC $0.23\!\to\!0.92$). The single measured regression
is truthfulness (TruthfulQA $-8.8$ points), a small and interpretable domain drift. A clean
vocabulary-expansion ablation ($<\!0.4$ pt on every general metric) confirms the gains are
attributable to CPT, not tokenizer changes. Crucially, subsequent SFT \emph{narrows} the model
back: all three axes fall to near-base levels, revealing a consistent division of labor ---
\textbf{CPT re-organizes and lifts the shared capability substrate; SFT cashes it out onto
target tasks}. We argue this reframes biological sequence not as a competitor for a foundation
model's capacity but as a form of structured scientific data that reshapes its capability
profile, and that CPT and SFT should be budgeted as complementary rather than substitutable
stages. We then confirm the mechanism causally: continuing pretraining of the same base seven
times under one fixed recipe, varying only the data mixture ($0/5/20/50\%$ biological share plus
a protein/DNA composition ablation), shows that raising the biological share to $50\%$ leaves the
general axis essentially unchanged (MMLU $-0.6$ pt, GSM8K $-1.2$ pt) while biological capability
rises monotonically (remote-homology MCC flips negative to positive; BixBench saturates), and
the \emph{type} of scientific data selects which capability is amplified. Mixture ratio is thus
a controllable, general-preserving lever on the capability profile. Finally, we ask the reverse
question at a scale with capability headroom the 26B model lacks: does biological CPT feed
anything back into ordinary language understanding? A matched-compute continuation experiment on
GPT-2 (124M), holding token budget, steps, and optimizer identical between a biology-continued
and a text-continued arm across three seeds, finds a real, reproducible transfer to the direct
structural counterpart of protein-homology detection --- paraphrase detection (PAWS-en,
$+3.1$ points, consistent in sign across all three seeds) --- alongside a genuine cost to
long-range discourse modeling (LAMBADA, $-14.6$ points), a partial rather than uniform
dissociation that we report honestly. All checkpoints, evaluation code, and per-example outputs
are public.
\end{abstract}

\medskip
\noindent\textbf{Keywords:} foundation models, continued pretraining, catastrophic forgetting,
data-centric AI, mixture-of-experts, capability transfer, structured scientific data.

% ===================================================================
\section{Introduction}
\label{sec:intro}


The dominant intuition about continued pretraining (CPT) on a narrow domain corpus is
subtractive: pouring out-of-distribution tokens --- raw protein and DNA sequence, which share
almost no surface statistics with natural language --- into a general-purpose model is expected
to erode its broad competence, a phenomenon variously described as the \emph{alignment tax} or
\emph{catastrophic forgetting} \citep{luo2023forgetting, ibrahim2024cpt}. Under this view,
domain adaptation is a trade: one buys biological ability at the cost of general ability, and
the practitioner's job is to minimize the damage.

This paper argues that, for a modern Mixture-of-Experts foundation model, the intuition is
wrong --- or at least incomplete. We show, by \emph{re-analysis of existing checkpoints and
without any new training}, that biological CPT on a 26B-parameter model does not merely avoid
catastrophic forgetting; it \emph{improves} the model on capability axes that have nothing to do
with biology, including general knowledge (MMLU~\citep{hendrycks2020measuring}) and code
generation (MBPP). We then show that
the subsequent supervised fine-tuning (SFT) stage, far from adding capability, \emph{narrows}
the model back toward its base competence on held-out axes. The picture that emerges is a clean
division of labor between the two training stages, and it suggests a reframing of what
biological data \emph{is} to a foundation model.

Our substrate is a Mixture-of-Experts model~\citep{shazeer2017sparse, fedus2022switch,
jiang2024mixtral} (Gemma-4-26B-A4B~\citep{gemmateam2025}); the analysis is deliberately
training-free, re-using an existing CPT/SFT lineage~\citep{ibrahim2024cpt} so that every
capability difference is attributable to a single stage. This complements router-level
interpretability of the same model family~\citep{wang2026omnigene, park2024moeinterpret,
kanai2025moeinterp} by asking not \emph{where} computation is organized but \emph{how} the
training data reshapes what the model can do.

\paragraph{Related work.}
Our claim sits at the intersection of three literatures. \emph{Continued pretraining and
forgetting}: domain-adaptive CPT is standard practice~\citep{ibrahim2024cpt, jindal2024balance,
jiang2024instruction}, but is widely reported to risk catastrophic forgetting of general
ability~\citep{luo2023forgetting, huang2024sssr}, motivating replay and balancing schemes. Most
of this evidence is on dense models and measures forgetting as a cost to be minimized; we instead
find a net \emph{gain} on a sparse MoE model and attribute it to the structural regularity of the
pretraining corpus rather than its factual content. \emph{Biological foundation models}: protein
and genomic models --- ESM-2~\citep{lin2023evolutionary}, ProtT5~\citep{elnaggar2021prottrans},
DNABERT-2~\citep{zhou2023dnabert2}, Nucleotide Transformer~\citep{dallatorre2023nucleotide}, the
Evo family~\citep{nguyen2024evo, brixi2026evo2}, and cross-modal LucaOne~\citep{he2024lucaone}, as
well as LLMs adapted to biology~\citep{luo2022biogpt, wang2026biopaws, wang2024llamagene} --- are
typically evaluated only on in-domain tasks, leaving open whether biological training helps or
harms general competence. \emph{Data-centric AI and scaling}: work on pretraining-data
composition and its effect on capability~\citep{gadre2021pile, hoffmann2022chinchilla} argues that
\emph{what} a model is trained on shapes \emph{what} it can do; we provide a controlled,
single-lineage instance of this principle where the manipulated variable is a scientific-sequence
corpus and the readout spans knowledge, code, and reasoning behavior. To our knowledge no prior
work reports that biological CPT \emph{improves} general and coding benchmarks, nor that it
measurably alters reasoning style.

\paragraph{Contributions.}
(i) A training-free re-analysis protocol over a single checkpoint lineage that isolates the
effect of biological CPT from tokenizer changes and from SFT. (ii) Evidence that biological CPT
lifts, rather than degrades, general and code capability, with a single interpretable regression
(truthfulness). (iii) A consistent CPT-lifts / SFT-narrows division of labor observed across four
independent evaluation axes. (iv) A data-centric reframing: biological sequence as structured
scientific data that reshapes a foundation model's capability profile. (v) A controlled
seven-model mixture experiment that turns the observational claim causal: varying only the CPT
data mixture, general ability is preserved across a $0\!-\!50\%$ biological share while biological
ability rises monotonically, and data \emph{composition} selects which capability is amplified.
(vi) A small-model, matched-compute causal probe of the reverse direction, showing that
biological CPT reproducibly improves natural-language paraphrase detection --- the direct
counterpart of the protein-homology transfer that motivates this line of work --- while honestly
reporting a real cost to long-range discourse and a scale at which character-level structural
probes are simply unmeasurable.

% ===================================================================
\section{Setup: one lineage, three checkpoints}
\label{sec:setup}
\input{sec_setup}

% ===================================================================
\section{Part I --- Observational re-analysis of one lineage}
\label{sec:results}
\input{sec_results}

% ===================================================================
\section{Part II --- Controlled mixture experiment}
\label{sec:phase2}
\input{sec_phase2}

% ===================================================================
\section{Part III --- A small-model, matched-compute causal probe}
\label{sec:phase3}
\input{sec_phase3}

% ===================================================================
\section{Discussion}
\label{sec:discussion}
\input{sec_discussion}

\section*{Data and code availability}
All three (four) checkpoints derive from a public lineage of Gemma-4-26B-A4B; the merged
BioCPT checkpoint is released as \texttt{dnagpt/OmniGene-4-CPT-v2-merged}. Evaluation scripts
(general, coding, biology-loglikelihood, and chain-of-thought diagnostics), the exact few-shot
configurations, and per-example model outputs for every table in this paper are released at
the project repository. All benchmarks (MMLU, ARC, HellaSwag, TruthfulQA, MBPP, HumanEval,
GSM8K, BixBench) are public; no proprietary data were used. The Part~I re-analysis is
training-free; the seven controlled Part~II CPT models ($100$M tokens each) together with their
QLoRA adapters, data-mixture recipes, and per-example evaluation outputs are also released at
the project repository. Part~II reproduces on a single 96\,GB GPU. The Part~III GPT-2 models
(one shared \texttt{nl\_base} plus six matched-compute continuation checkpoints, $3$ seeds
$\times$ 2 arms) are full-parameter finetunes released at the same repository together with the
character-manipulation probe data (mirrored from the original GPT-3-paper release) and reproduce
on a single consumer-grade GPU in under two hours total.

\section*{Acknowledgments}
% [placeholder]

\section*{Declaration of interests}
The author declares no competing interests.

\bibliographystyle{unsrtnat}
\bibliography{refs}

\end{document}
