
Our analysis rests on a single training lineage, so that every capability difference we
report can be attributed to a specific training stage rather than to confounds of architecture,
tokenizer, or data version.

\paragraph{Base model.} All checkpoints derive from Gemma-4-26B-A4B, a Mixture-of-Experts
transformer (128 experts per layer, top-8 routing, ${\approx}3.8$B active parameters). We use
the instruction-tuned release as the origin of the lineage; the biological CPT and SFT
checkpoints are those of OmniGene-4~\citep{wang2026omnigene}.

\paragraph{The three (four) checkpoints.} The lineage is
\[
\text{it} \;\xrightarrow{\text{vocab-expand}}\; \text{it-bio} \;\xrightarrow{\text{bio-CPT}}\;
\text{BioCPT} \;\xrightarrow{\text{SFT}}\; \text{BioCPT+SFT}.
\]
\textbf{(1) Base (it):} the original instruction-tuned model, vocabulary $262{,}144$.
\textbf{(2) Base (it-bio):} identical weights with the vocabulary expanded to $290{,}172$ by
adding $28{,}028$ biological BPE tokens (DNA/protein), but \emph{no further training}; this
serves as a clean control that isolates the cost of tokenizer expansion alone.
\textbf{(3) BioCPT:} it-bio after continued pretraining on $8.7$B tokens of mixed biological
corpus (DNA, protein sequence, and biomedical literature text), with LoRA ($r{=}64$) over
attention, MLP, and the MoE router, embeddings unfrozen; merged to full weights.
\textbf{(4) BioCPT+SFT:} BioCPT after supervised fine-tuning on biological instruction data
(a dual-head structural variant, ``v5''); merged to full weights.
We verified the lineage by direct weight comparison: it and it-bio share router/expert weights
bit-for-bit (mean abs.\ diff.\ $0$), differing only in embeddings, while BioCPT's router and
expert weights have measurably moved.

\paragraph{Evaluation.} We use the \texttt{lm-evaluation-harness} (v0.4.12) for all general and
coding benchmarks, at community-standard few-shot settings (MMLU 5-shot, ARC-Challenge 25-shot,
HellaSwag 10-shot, TruthfulQA-mc2 0-shot; HumanEval 0-shot, MBPP 3-shot). Multiple-choice tasks
are scored by loglikelihood, which is robust to a model's chat-vs-completion output format ---
a property we exploit deliberately, since our lineage mixes an instruction-tuned base with a
completion-style CPT checkpoint. Biomedical discrimination (BixBench) is likewise scored by
loglikelihood of the answer choices, and reasoning-behavior diagnostics (\S\ref{sec:results})
use a shared 8-shot chain-of-thought scaffold. All runs are on a single 96\,GB GPU in bfloat16;
the full model ($\approx$52\,GB) fits on one card, and the entire study required no training.

\paragraph{Honest scope.} Two evaluation regimes are format-sensitive and we flag them up front.
HumanEval 0-shot (bare function-completion) is unfair to instruction/SFT checkpoints and we
exclude it from conclusions, reporting MBPP 3-shot as the fair coding metric. Zero-shot protein
\emph{homology} is not measurable by any simple probe we tried (forced Yes/No loglikelihood is
near-random for all checkpoints; free generation is format-confounded); we therefore treat
sequence-homology ability separately and do not force it into the transfer analysis.
