
\subsection{A division of labor between CPT and SFT}
The recurring shape of our results --- base $<$ BioCPT $>$ SFT on every held-out axis --- points
to a division of labor that is usually conflated. Continued pretraining does not simply add
domain facts; it \emph{re-organizes and lifts a shared capability substrate} that is then
visible on tasks (general knowledge, code) far from the pretraining domain. Supervised
fine-tuning does the opposite: it \emph{narrows} the model onto the target task distribution,
trading breadth on held-out axes for focus on the served task. Neither stage is a substitute
for the other, and budgeting them as interchangeable --- or omitting CPT to ``save compute'' ---
misreads their roles. This is consistent with, and sharpens, prior router-level evidence on this
model family that CPT reshapes middle-layer computation while SFT re-aligns the output.

\subsection{Biological sequence as structured scientific data}
If narrow, out-of-distribution biological sequence \emph{raises} general and coding competence,
then the ``narrow domain = capacity competition'' framing is inadequate. We propose instead that
biological sequence functions as \emph{structured scientific data}: a corpus whose value to a
foundation model lies not (only) in the facts it carries but in the structural regularities it
forces the model to internalize --- regularities that transfer to other structured/formal
languages such as code. Under this view biology is simply the first representative member of a
broader class (code, mathematics, chemistry/SMILES), and the natural next question is not
``does biology help biology'' but ``how does the \emph{composition} of scientific-data CPT
shape the capability profile of a foundation model.''

\subsection{Mixture ratio is a controllable, general-preserving lever}
The controlled sweep of \S\ref{sec:phase2} answers that question for this model. The two axes do
not trade off: across a biological share of $0\!-\!50\%$ the general axis is flat while the
biology axis rises monotonically, so there is no ``forgetting threshold'' to budget around
within this range. Equally, the composition ablation shows the lever has structure --- DNA and
protein amplify \emph{different} biological capabilities --- so a practitioner can, in principle,
target a desired capability profile by choosing not just how much scientific data to add but
which kind. This is the operational form of the data-centric thesis: training-data composition
is a design surface for the capability profile, not merely a source of facts.

\subsection{Capability shaping does not require routing reorganization}
The router analysis of \S\ref{sec:router} complicates a tempting mechanistic story. It would be
neat if the mixture experiment's capability dose-response were simply the LoRA-scale echo of the
routing reorganization we observe in the full-CPT lineage --- more biology, more specialized
experts, more biology capability. The data do not support that story: routing divergence in the
mixture sweep is flat with respect to biological share even as capability rises monotonically,
and a purely \emph{text-only} $100$M-token LoRA run already produces most of the routing shift
that biological runs do. What this suggests instead is that a modest LoRA-scale CPT run shifts
\emph{how} the model routes tokens in a largely data-agnostic way (any continued pretraining
perturbs routing by a similar amount), while \emph{what capability that shift buys} still
depends closely on what the model was shown. Capability and routing are therefore separable
readouts of continued pretraining, not two views of the same underlying change, and only
converge once the training budget is large enough (here, $8.7$B full-parameter tokens) to
substantially reorganize the router itself. This is a useful clarification for anyone using
routing metrics as a proxy for capability change under LoRA-scale adaptation: at this scale, the
proxy does not track the target.

\subsection{The transfer is real, but scale-dependent and narrow}
Parts I and II both concern a 26B model whose general axis is near ceiling on the benchmarks we
use; ``preserved, not improved'' is the honest description of what data composition buys there.
Part III asks the sharper causal question --- does biological sequence \emph{add} natural-
language capability, not merely leave it undisturbed --- in a regime with headroom, and finds a
real but narrow answer: paraphrase detection improves reproducibly, several other structural
probes are unmeasurable at this scale, and a real cost appears in long-range discourse. This
resolves an apparent tension rather than creating one. The forward-transfer hypothesis that
motivates this paper --- a shared difference-detection structure between natural-language
paraphrase and sequence homology --- predicts a \emph{specific} transfer target, not a general
uplift of all natural-language skill; PAWS-en is exactly that target, and it is the one metric
that moves. That the effect appears only where capacity is scarce (a $124$M full finetune) and
not where it is abundant (a $26$B MoE with LoRA-isolated updates) is itself informative: it
suggests biological CPT's natural-language footprint is a genuine but small-magnitude structural
signal that a large, high-capacity model can absorb without displacing anything, while a small,
fully-tuned model trades it against other capability. Data composition, in other words, shapes
capability at every scale we tested, but what counts as a ``free'' addition versus a
``competed-for'' resource depends on how much spare capacity the training regime leaves it.

\subsection{Limitations}
This is a training-free re-analysis of one lineage of one 26B MoE model; the CPT / SFT recipe is
fixed, so we characterize \emph{this} lineage rather than establishing a scaling law. Part of the
ARC/HellaSwag gain is a format effect (instruct-vs-completion loglikelihood), which we flag and
which does not affect the knowledge-heavy MMLU gain. Zero-shot sequence-homology is not
measurable by simple probes, so that axis is argued indirectly. TruthfulQA is a genuine, if
small, regression and we do not hide it. The controlled experiment of \S\ref{sec:phase2}
addresses the observational limitation directly --- it fixes the recipe and varies only the data
mixture --- but was itself bounded: each model saw $100$M CPT tokens (a pilot budget chosen to
find trends, not a scaling law), on one 26B MoE model. \S\ref{sec:scaling} closes the model-size
gap (the general-preserving pattern holds from $2.3$B to $26$B) but not the token-budget gap:
whether the pattern holds at $10\times$ the tokens, or for other scientific-data domains (code,
mathematics, chemistry), remains open. Part III is deliberately narrow: one small model
(GPT-2, $124$M), one continuation recipe, and a character-level probe battery that is
unmeasurable at this scale in either arm, so we can report only that PAWS-en transfers, not that
the full structural family does; whether the effect grows, saturates, or reverses at
intermediate scales between $124$M and $2.3$B (our smallest Part IV model) remains open. We did
test whether the long-range-discourse cost of Part II's uniform mixing could be mitigated by an
explicit replay schedule (\S\ref{sec:replay}) --- it could not, because there was nothing to
mitigate at this budget --- but did not test whether a \emph{lower} biological share changes
that picture, nor whether Part III's full-parameter, single-model finding generalizes across
seeds and model families the way Part II's LoRA finding does across three sizes. Finally, the
router analysis of \S\ref{sec:router} covers only this one model family and MoE configuration;
whether the LoRA/full-CPT routing dissociation is a general property of MoE continued
pretraining or specific to this architecture and token regime is a question for future work.
